---
title: "Uber Data Analysis"
author: "Derrick"
date: "2023-11-06"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com"))
```

### Uber Dataset

#### Details of the dataset:

##### The dataset contains information about the Datetime, Latitude, Longitude and Base of each uber ride that happened in the months of april, may, june, july, august and september 2014 at New York City, USA.

##### Date/Time : The date and time of the Uber pickup

##### Lat : The latitude of the Uber pickup

##### Lon : The longitude of the Uber pickup

##### Base : The TLC base company code affiliated with the Uber pickup

##### The Base codes are for the following Uber bases:

##### B02512 : Unter

##### B02598 : Hinter

##### B02617 : Weiter

##### B02682 : Schmecken

##### B02764 : Danach-NY

```{r Loading the packages}
install.packages("tidyverse")
install.packages("pillar")
install.packages("dplyr")
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(stringr)
library(readr)
```

```{r importing the data}
apr <- read_csv("uber-raw-data-apr14.csv")
may <- read_csv("uber-raw-data-may14.csv")
jun <- read_csv("uber-raw-data-jun14.csv")
jul <- read_csv("uber-raw-data-jul14.csv")
aug <- read_csv("uber-raw-data-aug14.csv")
sep <- read_csv("uber-raw-data-sep14.csv")
```

```{r Exploring the Data}
# explore the structure of the dataset
str(apr)
str(may)
str(jun) 
str(jul)
str(aug)
str(sep)
# from this, it is easy to combine the data
# this is because they all have the same structure
# Hence the data can be combined to form one large set

# Combining the Data
cdata <- bind_rows(apr, may, jun, jul, aug, sep)
head(cdata)
tail(cdata)
# Summarizing the combined data
summary(cdata)
```

```{r Handling missing values}
# Identify the missing values
missing_values <- colSums(is.na(cdata))
# View columns with missing values
print(missing_values[missing_values > 0])

# There are no missing values in this dataset
```

```{r Handling Duplicate Data}
# Identify and remove duplicate rows
cdata <- distinct(cdata)
```

```{r feature engineering}
# Load the lubridate package
library(lubridate)

# Assuming your data frame is named 'cdata' and has a column 'Date/Time'
cdata$`Date/Time` <- as.POSIXct(cdata$`Date/Time`, format = "%m/%d/%Y %H:%M:%S", tz = "UTC")

# Assuming 'data' is your data frame
cdata$Date <- as.Date(cdata$`Date/Time`)

# Create new column for Time
cdata$Time <- format(cdata$`Date/Time`, format = "%H:%M:%S")
head(cdata)
```

```{r more engineering}
# Create individual columns for month day and year
cdata$day <- factor(day(cdata$`Date/Time`))
cdata$month <- factor(month(cdata$`Date/Time`, label=TRUE))
cdata$year <- factor(year(cdata$`Date/Time`))
cdata$dayofweek <- factor(wday(cdata$`Date/Time`, label=TRUE))

# Add Time variables as well 
cdata$second = factor(second(hms(cdata$`Time`)))
cdata$minute = factor(minute(hms(cdata$`Time`)))
cdata$hour = factor(hour(hms(cdata$`Time`)))

# Look at the data
head(cdata)
```

```{r column arrange}
cdata <- cdata[, c("Lat","Lon", "Base", "Date/Time", "Date", "Time", "second","minute","hour", "day", "dayofweek", "month", "year")]
head(cdata)
tail(cdata)
```

```{r Saving cleaned dataset}
write.csv(cdata, "cleaned_dataset.csv", row.names = FALSE)
```

## Visualizations

```{r data aggregation}
hourly_data <- cdata %>% 
                    group_by(hour) %>% 
                            dplyr::summarize(Total = n())
# Plot the data by hour
ggplot(hourly_data, aes(x = hour, y = Total)) + geom_bar(stat="identity", fill="lightblue", color="maroon") + 
ggtitle("Trips made per Hour") + 
theme(legend.position = "none", 
      plot.title = element_text(hjust = 0.5), 
      plot.subtitle = element_text(hjust = 5))
```

#### The highest pickup hour is 17:00, mostly when people are leaving from their working places.

```{r month}
day_data <- cdata %>% 
                    group_by(day) %>% 
                            dplyr::summarize(Trips = n())

# Plot the data by day
ggplot(day_data, aes(day, Trips)) + geom_bar(stat="identity", fill="lightblue", color="maroon") + 
ggtitle("Trips made per day") + 
theme(legend.position = "none", 
      plot.title = element_text(hjust = 0.5), 
      plot.subtitle = element_text(hjust = 5))
```

```{r pie chart}
# Aggregate the data by month and hour
month_hour_data <- cdata %>% 
  group_by(month, hour) %>%  
  summarise(Total = n())

# Create a pie chart using ggplot2
piechart <- ggplot(month_hour_data, aes(x = "", y = Total, fill = month)) + 
  geom_bar(stat = "identity",width = 0.5, color = "white") + 
  coord_polar("y") +
  ggtitle("Trips by Hour and Month") +
  theme_void()

# Display the pie chart
print(piechart)
```

#### The month with the highest trips made was september, followed by august, july, june, may and lastly april.

```{r day of week}
# Collect data by day of the week and month

day_month_data <- cdata %>% 
  group_by(dayofweek, month) %>% 
  summarize(Trips = n())
day_month_data

# Assuming colors is a vector of 6 colors
colors <- c("blue", "darkgreen", "yellow", "brown", "red", "maroon")

# Plot the data
ggplot(day_month_data, aes(dayofweek, Trips, fill = month)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  ggtitle("Trips by Day and Month") + 
  scale_fill_manual(values = colors)
```

## Creating a map visualization of rides in NYC

```{r Map to show distribution}
# Set min and max lat/long
min_lat <- 40
max_lat <- 40.91
min_long <- -74.15
max_long <- -73.7004

# Create the plot
ggplot(cdata, aes(x = Lon, y = Lat)) +
  geom_point(size = 1, color = "green") +
  scale_x_continuous(limits = c(min_long, max_long)) +
  scale_y_continuous(limits = c(min_lat, max_lat)) +
  labs(title = "NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP)")
```
